---
title: "ps05-rmarkdown-plot"
author: "Zack Tisdale"
date: "2023-02-17"
output: html_document
---
```{r, echo = FALSE}
library(tidyverse)
```
## 1. Load and Check Data (5pts)
The data loading in: 
```{r}
gapMinder <- read_delim("data/gapminder.csv.bz2")
```
The dimensions are as follows `r dim(gapMinder)`

A sample of the data is as follows
```{r}
head(gapMinder,5)

```
It doesn't look too good

## 2. descriptive Statistics (15pts)

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.


Country count in iso3
```{r}
gapMinder %>% 
  select(iso3) %>% 
  na.omit() %>% 
  summarize(n = n_distinct(iso3))
```
Country count in iso2
```{r}
gapMinder %>% 
  select(iso2) %>% 
  na.omit() %>% 
  summarize(n = n_distinct(iso2))
```
Country count in name
```{r}
gapMinder %>%
  select(name) %>% 
  na.omit() %>% 
  summarize(n = n_distinct(name))
```

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?

Names
```{r, error = TRUE}
gapMinder%>% 
   filter(is.na(name)) %>% 
   distinct(iso2)
```

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
gapMinder%>% 
   filter(is.na(name)) %>% 
   distinct(iso3)
```
3. (2pt) What is the minimum and maximum year in these data?
min
```{r}
gapMinder %>% 
  select(time) %>% 
  na.omit() %>% 
  min()
```
max
```{r}
gapMinder %>% 
  select(time) %>%
  na.omit %>% 
  max()
```


## 3 Co2 emissions (30pt)

Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

Missing Co2 emissions stats for each year
```{r}
gapMinder %>% 
  group_by(time) %>% 
  select(co2,time) %>% 
  summarize(n = sum(is.na(co2)))
```
Missing Co2_PC emissions stats for each year
```{r}
gapMinder %>% 
  group_by(time) %>% 
  select(co2_PC,time) %>% 
  summarize(n = sum(is.na(co2_PC)))
```
Top 5 years with the most missing data for Co2
```{r}
gapMinder %>% 
  group_by(time) %>% 
  select(co2,time) %>% 
  summarize(n = sum(is.na(co2))) %>% 
  arrange(desc(n)) %>% 
  head(5)
```
Top 5 years with the most missing data for co2_PC
```{r}
gapMinder %>% 
  group_by(time) %>% 
  select(co2_PC,time) %>% 
  summarize(n = sum(is.na(co2_PC))) %>% 
  arrange(desc(n)) %>% 
  head(5)
```
2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r}
miscCountries <- gapMinder %>% 
  filter(name == 'United States of America'|name == 'China' | name == 'India'| name == 'Japan' | name == 'Russian Federation')
ggplot(miscCountries) +
  geom_point(aes(x = time, y = co2, col = name))+
  ggtitle("Co2 Emissions Among Countries")+
  xlab("Year") +
  ylab("Emissions (kt)")
```
It seems that China is an anomoly of the countries, although the the rest of the countries except the US seem to have a slow and steady growth
3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
ggplot(miscCountries) +
  geom_point(aes(x = time, y = co2_PC, col = name))+
  ggtitle("Co2 Emissions per Captia Among Countries")+
  xlab("Year") +
  ylab("Emissions per Capita (kt)")
```
This suggests that the US has a lot more Co2 per person compared to other countries, and that china and india have a lot less co2 in the graph when you scale it by popuation

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
```{r} 
continents <- gapMinder %>% 
  select(time,region,co2_PC) %>% 
  na.omit() %>% 
  group_by(time,region) %>%
  summarize(n = mean(co2_PC))
ggplot(continents) +
  geom_point(aes(x = time, y = n, col = region)) +
  ggtitle("Average co2 emissions per Capita across Continents")+
  xlab("year")+
  ylab("Mean emssions per capita (metric tons)")

```

I see that the Americas, and Asia wre all over the charts near the beginning, and 
they started to level out more so after the 90s

Hint: Americas 2016 should be 4.80.
5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

#### 3.4 attempt 2
```{r}
continents <- gapMinder %>%
  filter(time == 1960 | time == 2016) %>% 
  select(region,co2_PC,time) %>% 
  na.omit() %>% 
  group_by(time,region) %>% 
  summarize(n = mean(co2_PC))
continents
ggplot(continents)+
  geom_col(aes(x = region, y = n, fill=factor(time)),position = "dodge") +
  ggtitle("Average CO2 per capita per Continent in 1960 and 2016")+
  xlab("Continents")+
  ylab("Mean CO2 per captia (metric tons)")
```


Hint: it should look something along these lines:
## GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC ).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r}
countries1960 <- gapMinder %>% 
  select(GDP_PC, time, lifeExpectancy, totalPopulation, region) %>% 
  na.omit() %>% 
  filter(time == 1960)

ggplot(countries1960) +
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col = region),
             alpha = 0.7) +
  xlab("GDP per capita") +
  ylab("Life Expectancy")+
  ggtitle("GDP per captia vs Life Expectancy in 1960")
```

2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
countries2019 <- gapMinder %>% 
  select(GDP_PC, time, lifeExpectancy, totalPopulation, region) %>% 
  na.omit() %>% 
  filter(time == 2019)

ggplot(countries2019) +
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col = region),
             alpha = 0.7) +
  xlab("GDP per capita") +
  ylab("Life Expectancy")+
  ggtitle("GDP per captia vs Life Expectancy in 2019")
```

They fit
3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

The life expectancy has congregated a lot higher up, the populations have grown a lot 
as well, and there seems to be a lot more countries with a higher gdp. Also centers 
of the graphs are farther right and higher.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.
```{r}
lifeExpectancyTimes <- gapMinder %>% 
  filter(time == 1960 | time == 2019) %>% 
  group_by(region,time) %>% 
  select(region,time,lifeExpectancy) %>% 
  na.omit() %>% 
  summarize(n = mean(lifeExpectancy))
ggplot(lifeExpectancyTimes)+
  geom_col(aes(x = region, y = n, fill=factor(time)),position = "dodge") +
  ggtitle("Average Life Expectancy per Continent in 1960 and 2016")+
  xlab("Continents")+
  ylab("Mean Life Expectancy")
```



I think the results fit with the figures that were created


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r}
LEGrowth <- gapMinder %>% 
  filter(time == 1960| time == 2019) %>% 
  group_by(region, time) %>%
  select(region,time,lifeExpectancy) %>% 
  na.omit %>% 
  summarize(n = mean(lifeExpectancy)) %>% 
  mutate(prev = lag(n),
         growth = n - prev) %>% 
  na.omit() %>% 
  arrange(desc(growth)) %>% 
  select(region,growth)
LEGrowth
```
I see that the order of the biggest growth, also that the continents that had the
largest highest life expectancy increased the it seems.


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
gpdPerCaptia <- gapMinder %>% 
  filter(time == 1960| time == 2019) %>%
  group_by(time) %>% 
  select(GDP_PC,time) %>% 
  na.omit()
ggplot(gpdPerCaptia) +
  geom_histogram(aes(x = GDP_PC, fill = factor(time)),position = "dodge") +
  ggtitle("GDP per captia")+
  xlab("GDP per Capita") +
  ylab("Number of Countries")

```
7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}
usLifeExpectancyRank <- gapMinder %>% 
  group_by(time) %>% 
  select(name,time,lifeExpectancy) %>% 
  na.omit %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == 'United States of America') %>% 
  select(rank,time)
ggplot(usLifeExpectancyRank)+
  geom_line(aes(x = time, y = rank)) + 
  ggtitle("US rank in Life Expectancy over time") +
  xlab("Year") +
  ylab("Rank Number")
```
8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a 
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.

```{r}
usLifeRelativeExpectancyRank <- gapMinder %>% 
  group_by(time) %>% 
  select(name,time,lifeExpectancy) %>% 
  na.omit %>% 
  mutate(rank = rank(desc(lifeExpectancy))/n_distinct(name)) %>% 
  filter(name == 'United States of America') %>% 
  select(rank,time)
usLifeRelativeExpectancyRank
ggplot(usLifeRelativeExpectancyRank)+
  geom_line(aes(x = time, y = rank)) + 
  ggtitle("US relative rank in Life Expectancy over time") +
  xlab("Year") +
  ylab("Rank Number divided by the number of countries in that year in the data")
```

Finally tell us how many hours did you spend on this PS.

I would say around 7 or 8 at least